---
title: "dfafda"
output: ioslides_presentation
---


# Building and Using Ensembl-based Annotation Packages with `ensembldb`<a id="orgheadline17"></a>

## Introduction<a id="orgheadline1"></a>

-   `TxDb` objects from `GenomicFeatures` provide gene model annotations.
-   `ensembldb` package defines `EnsDb` class also providing gene models.
-   `EnsDB` object/database:
    -   Designed for Ensembld based annotations: new attributes *gene biotype* and *tx biotype.*
    -   Allow to query specific annotations using a simple filter framework.
    -   Methods allow to specify the type of the returned object.
    -   Interchangeable with `TxDb` objects.
    -   Support `AnnotationDbi`'s `select` method.

## Usage<a id="orgheadline9"></a>

### Query gene, transcript, exon information<a id="orgheadline2"></a>

-   `EnsDb` objects can be used just like `TxDb` objects.
-   Available methods:
    -   `genes`
    -   `transcripts`
    -   `transcriptsBy`
    -   `exons`
    -   `exonsBy`
    -   `cdsBy`
    -   `fiveUTRsByTranscripts`
    -   `threeUTRsByTranscripts`
-   Additions:
    -   Define the type of object we want to get.
    -   Use of filter(s) to query specific information.
-   <span class="underline">Example</span>: get genes and transcript models.

```{r }
## Load an EnsDb package matching Ensembl version 81
library(EnsDb.Hsapiens.v81)
edb <- EnsDb.Hsapiens.v81

## List the available tables and columns
listTables(edb)

## Now just get all genes
genes(edb)

## Alternatively, we can specify to return the results as DataFrame
transcripts(edb, return.type="DataFrame")
```

-   <span class="underline">Example</span>: Use filters to fetch specific information.

```{r }
####
## Retrieve genes encoded on chromosome Y.
## Create a filter object
sf <- SeqnameFilter("Y")

sf

## Retrieve the data
genes(edb, filter=SeqnameFilter("Y"))

####
## Combine filters:
## Create a 'gene biotype' filter
listGenebiotypes(edb)

## Define a filter that returns everything except protein coding genes
bf <- GenebiotypeFilter("protein_coding", condition="!=")

bf

## Get all non-coding genes from chromosome Y
gns <- genes(edb, filter=list(sf, bf))

gns

sort(table(gns$gene_biotype))
```

### Supported filter objects.<a id="orgheadline3"></a>

-   For genes:
    -   `GeneidFilter`
    -   `GenenameFilter`
    -   `EntrezidFilter`
    -   `GenebiotypeFilter`
-   For transcripts:
    -   `TxidFilter`
    -   `TxbiotypeFilter`
-   For Exons:
    -   `ExonidFilter`
    -   `ExonrankFilter`
-   *Generic* filters:
    -   `SeqnameFilter`
    -   `SeqstrandFilter`
    -   `SeqstartFilter`
    -   `SeqendFilter`
    -   `GRangesFilter`: condition can be *within* or *overlapping*.

-   Multiple filters can be combined with a logical AND.
-   <span class="underline">Example</span>: combine filters.

```{r }
####
##  Example for a GRangesFilter
##  condition could be "within" or "overlapping"
grf <- GRangesFilter(GRanges(17, IRanges(59000000, 59200000)),
		     condition="within")
grf

genes(edb, filter=grf)

####
## Combine filters
## Get only miRNA genes.
genes(edb, filter=list(grf, GenebiotypeFilter("miRNA")))


####
## Other example:
##
## Get all first exons of all genes encoded on chromosome X
exons(edb, filter=list(ExonrankFilter(1), SeqnameFilter("X")),
      columns=c("tx_id", "gene_name", "exon_id", "exon_idx"))
```

### Annotation for feature counting<a id="orgheadline4"></a>

-   In combination with `summarizeOverlaps` from the `GenomicAlignments` package.
-   <span class="underline">Example</span>: get gene models for feature counting with `summarizeOverlaps`.

```{r eval=FALSE}
####
## Count reads within exons per gene:
## -> get the exons by gene, for chromosome 1:22, X, Y,
##    excluding also locus reference genomic genes (LRG)
exns <- exonsBy(edb, by="gene", filter=list(SeqnameFilter(c(1:22, "X", "Y")),
					    GeneidFilter("ENSG%", "like")))
exns

## Load the required libraries.
library(GenomicAlignments)
library(BiocParallel)

## Get the Bam files.
bfl <- BamFileList(dir("data/bam", pattern=".bam$", full.names=TRUE),
		   asMates=TRUE, yieldSize=1e+6, obeyQname=TRUE)
## Define a ScanBamParam with a mapping quality filter.
sbp <- ScanBamParam(mapqFilter=30)

##### DONTRUN
## Do the gene counting
geneCounts <- bplapply(bfl, FUN=summarizeOverlaps, features=exns,
		       mode="IntersectionStrict", ignore.strand=TRUE,
		       singleEnd=FALSE, fragments=TRUE, param=sbp)
geneCounts <- do.call(cbind, geneCounts)
####

## Load pre-calculated counts
load("data/RData/tmp-geneCounts-all-chrom.RData")

head(assay(geneCounts))
```

-   In combination with `Rsubread`'s `featureCount` function.
-   <span class="underline">Example</span>: gene models for `featureCount`.

```{r eval=FALSE}
## Convert the exon list to SAF format
saf <- toSAF(exns)

head(saf)

####
##  Do the feature counting using the Rsubread package
library(Rsubread)
bamf <- dir("data/bam", pattern=".bam$", full.names=TRUE)
cnts <- featureCounts(files=bamf, annot.ext=saf, isPairedEnd=TRUE, nthreads=1)
```

### Get genomic sequences etc.<a id="orgheadline5"></a>

-   `BSGenome` packages provide genomic sequence for a variety of species.
-   Most are based on UCSC data, that use different genome
    release names than Ensembl does (e.g. *hg19* instead of *GRCh37*).
-   <span class="underline">Example</span>: get mRNA sequences for a gene's transcripts.

```{r }
####
##  Get the genome version of the package:
unique(genome(edb))

## Where to get this genome sequence?
## Answer: AnnotationHub
library(AnnotationHub)
ah <- AnnotationHub()
## Search for all human Ensembl files for release 81.
query(ah, pattern=c("Ensembl", "homo sapiens", "81"))

## So we've got the genomic DNA sequence matching that version.
fafile <- ah[["AH49186"]]

## Easier way: use the dedicated getGenomeFaFile method
fafile <- getGenomeFaFile(edb)

## Get all exons by transcript.
ska2Txs <- exonsBy(edb, by="tx", filter=GenenameFilter("SKA2"))

## Get sequence of individual exons.
getSeq(fafile, ska2Txs[[1]])

## Alternative: use extractTranscriptSeqs from GenomicFeatures.
txSeqs <- extractTranscriptSeqs(fafile, ska2Txs)
txSeqs
```

-   That's convenient, especially if we're using an `EnsDb` for a non-standard
    organism.
-   **But**: wouldn't it be nice to use UCSC genome sequences too?
-   <span class="underline">Example</span>: use `EnsDb` with UCSC style chromosome names.

```{r }
####
##  Try to get the sequence from the BSgenome package.
library(BSgenome.Hsapiens.UCSC.hg38)

## Doesn't work, because of different chromosom naming.
##extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38, ska2Txs)

## Solution: change the seqlevel style
seqlevelsStyle(edb)
seqlevelsStyle(edb) <- "UCSC"

## Now we can use UCSC chromosome names:
genes(edb, filter=SeqnameFilter("chrY"))

ska2Txs <- exonsBy(edb, filter=GenenameFilter("SKA2"))
seqlevels(ska2Txs)

extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38, ska2Txs)

## Compare to the FaFile from Ensembl.
txSeqs

## Change back, so we don't break following examples.
seqlevelsStyle(edb) <- "Ensembl"
```

-   Sequence names are mapped between *styles* using the `GenomeInfoDb` package.

### Plotting<a id="orgheadline6"></a>

-   `Gviz`: `getGeneRegionTrackForGviz` method to extract a `data.frame` formatted for
    `Gviz`.
-   <span class="underline">Example</span>: plot genes encoded on a chromosomal region using `Gviz`.

```{r }
library(Gviz)

## Get the gene SKA2
ska2 <- genes(edb, filter=GenenameFilter("SKA2"))

grt <- getGeneRegionTrackForGviz(edb, chromosome=seqlevels(ska2),
				 start=start(ska2), end=end(ska2))
geneTrack <- GeneRegionTrack(grt)
theTracks <- list(geneTrack)
plotTracks(theTracks, transcriptAnnotation="symbol",
	   chromosome=seqlevels(ska2))

## BAM_PATH <- "data/bam/"

## ##bamF <- c("A-chr5-17.bam", "B-chr5-17.bam")
## bamF <- c("A.bam", "B.bam")
## algnTracks <- lapply(bamF, function(z){
##     return(AlignmentsTrack(paste0(BAM_PATH, z), name=z))
## })

## gat <- GenomeAxisTrack()
## GRCh38 <- getGenomeFaFile(edb)
## SI <- seqinfo(GRCh38)[c(1:22, "X", "Y")]
## gr <- GRanges(seqnames=seqnames(SI), IRanges(rep(1, length(SI)),
##                                              seqlengths(SI)))
## ## Get the sequences,
## dnas <- getSeq(GRCh38, gr)
## seqTr <- SequenceTrack(dnas, chromosome="1", genome="GRCh38", name="GRCh38")

## ska2 <- genes(edb, filter=GenenameFilter("SKA2"))

## grt <- getGeneRegionTrackForGviz(edb, chromosome=seqlevels(ska2),
##                                  start=start(ska2), end=end(ska2))
## geneTrack <- GeneRegionTrack(grt)

## theTracks <- list(gat, seqTr, geneTrack)
## plotTracks(theTracks, transcriptAnnotation="symbol",
##            chromosome=seqlevels(ska2))

## plotTracks(c(theTracks, algnTracks), transcriptAnnotation="symbol",
##            chromosome=seqlevels(ska2))
## ??? not plotting the alignment?
```

-   `ggbio`: we can directly pass `EnsDb` objects and filters to the plotting
    function.
-   <span class="underline">Example</span>: use `ggbio` and `ensembldb` to plot a chromosomal region.

```{r }
## Use ggbio:
library(ggbio)

## Plot the SKA2 gene model.
autoplot(edb, GenenameFilter("SKA2"))

## Or get all genes from the SKA2 genomic locus.
ska2ChromReg <- GRanges(seqnames=seqlevels(ska2),
			IRanges(start(ska2), end(ska2)))

## Plot the genomic region; we're using the gene name as labels for the transcripts.
autoplot(edb, GRangesFilter(ska2ChromReg), names.expr="gene_name")
```

### `ensembldb` and the `AnnotationDbi` API<a id="orgheadline7"></a>

-   `AnnotationDbi` defines methods `columns`, `keys`, `keytypes`, `mapIds` and `select`.
-   `EnsDb` support all these methods which can be combined with *filters*.
-   <span class="underline">Example</span>: use methods from the `AnnotationDbi` API to fetch annotations.

```{r }
## Get the supported 'columns' from an EnsDb
columns(edb)

## Different from listColumns
listColumns(edb)

## What keytypes can be used
keytypes(edb)

## List all tx keys (tx_id)
head(keys(edb, keytype="TXID"))

## Note that we can again use filters here:
txKeys <- keys(edb, keytype="TXID", filter=GenenameFilter("SKA2"))
txKeys

## Use these keys in the select call
select(edb, keys=txKeys, keytype="TXID")

## Or use the filter directly in the select call: submit filters with
## 'keys' argument.
select(edb, keys=GenenameFilter("SKA2"))
```

### The `ensembldb` *shiny* app<a id="orgheadline8"></a>

-   A `shiny` app is also available for `EnsDb` objects, start it with `runEnsDbApp`.
-   <span class="underline">Example</span>: search for a gene using the shiny app and return the result to R.

```{r eval=FALSE}
## Run the shiny app:

Result <- runEnsDbApp()

## Inspect the result:
Result
```

## Building annotation databases<a id="orgheadline13"></a>

-   Ensembl, Ensembl Genomes and Ensembl Plants provide annotations for a large
    number of species. All follow the same database scheme.

### The *hard way*: with Ensembl's Perl API<a id="orgheadline10"></a>

-   Requires:
    -   Ensembl Perl API (and Bioperl).
    -   `PERL5LIB` environment variable should point to the API version matching the
        Ensembl version that should be queried.
-   `fetchTablesFromEnsembl` to fetch the annotations from Ensembl.
-   `makeEnsemblSQLiteFromTables` to create the SQLite database from the tables.
-   `makeEnsembldbPackage` to create a package containing and providing the
    annotation.
-   <span class="underline">Example</span>: create an `EnsDb` using the Perl API.

```{r eval=FALSE}
## Example to create an EnsDb using the Ensembl Perl API:

## This takes quite some time...
fetchTablesFromEnsembl(version="81",
		       ensemblapi="/Users/jo/ensembl/81/API/ensembl/modules",
		       species="dog")

## Create an SQLite database from the generated txt files
dbf <- makeEnsemblSQLiteFromTables()

## Finally, create the package
makeEnsembldbPackage(ensdb=dbf, version="1.0.0",
		     maintainer="Johannes Rainer <johannes.rainer@eurac.edu>",
		     author="Johannes Rainer")
```

### The *easy way*: from gtf and gff files<a id="orgheadline11"></a>

-   Easier and faster way: create an `EnsDb` from a *gtf* or *gff* file.
-   *Should* work with all gtf and gff files from Ensembl.
-   Gtf files don't provide the Entrez IDs, database column will thus be empty.
-   <span class="underline">Example</span>: create an `EnsDb` from a GTF file downloaded from <ftp://ftp.ensembl.org>.

```{r }
## Create a EnsDb from a GTF file from Ensembl.

## Create the SQLite database file:
##  o Eventually define 'organism' and 'genomeVersion'.
##  o Needs also an internet connection to retrieve the 'seqlengths'.
edbSql <- ensDbFromGtf("data/gtf/Canis_familiaris.CanFam3.1.84.gtf.gz")

edbSql

## Use the makeEnsembldbPackage to create a package, or load and use it.
dogDb <- EnsDb(edbSql)

dogDb

## Fully functional, except we don't have Entrez gene ids.
head(genes(dogDb, filter=SeqnameFilter("X")))
```

### Even easier: use `AnnotationHub`<a id="orgheadline12"></a>

-   We can build an `EnsDb` from an `AnnotationHub` resource or a `GRanges` object.
-   <span class="underline">Example</span>: create an `EnsDb` using `AnnotationHub`.

```{r }
## Create an EnsDb from a GRanges retrieved from AnnotationHub.

library(AnnotationHub)
ah <- AnnotationHub()

## Query for available Ensembl files matching homo sapiens and 83
query(ah, pattern=c("ensembl", "83", "canis"))

## Get the GRanges object matching the version 83 gtf.
edbSql83 <- ensDbFromAH(ah=ah["AH50358"])

## Make a package from that.
makeEnsembldbPackage(ensdb=edbSql83, version="1.0.0",
		     maintainer="Johannes Rainer <johannes.rainer@eurac.edu>",
		     author="J Rainer")



## Alternatively, we could build the EnsDb from a GRanges object
## providing the required columns/attributes.
ensGr <- ah[["AH50358"]]

ensGr

## Create a EnsDb from that GRanges
##  o have to provide organism, genome version and (Ensembl) version
edbSql83 <- ensDbFromGRanges(ensGr, organism="Canis familiaris", version="83",
			     genomeVersion="CanFam3.1")
```

## Some internals on the filter framework<a id="orgheadline14"></a>

-   How does the filter framework work?
-   <span class="underline">Example</span>: `column` and `where` methods of a `GenenameFilter`.

```{r }
## Create a simple GenenameFilter
gnf <- GenenameFilter("SKA2")

gnf

## 'column' and 'where' methods.
column(gnf)
where(gnf)

## These are conditional on the database object
column(gnf, edb)
where(gnf, edb)
```

-   `column` method is used to define which columns are required from the database.
-   `where` method is used to build the *where* condition of the SQL call.
-   Implementing these methods for other database objects enables reuse of the
    classes in other packages.

## Other packages using the `ensembldb` filter framework<a id="orgheadline15"></a>

-   `mirhostgenes` R-package (<https://github.com/jotsetung/mirhostgenes>): predicts
    host genes (primary transcription units) of miRNAs based on genomic
    alignment of pre-miRNAs.
-   Imports all filter objects from `ensembldb`.
-   <span class="underline">Example</span>: get all miRNAs potentially encoded in the same primary transcript
    than SKA2.

```{r }
## Plot the genomic region in which SKA2 is encoded to illustrate the
## host gene prediction.
library(Gviz)
ska2 <- genes(edb, filter=gnf)
chromReg <- getGeneRegionTrackForGviz(edb, chromosome=seqlevels(ska2),
				      start=start(ska2), end=end(ska2))
geneTrack <- GeneRegionTrack(chromReg)
plotTracks(geneTrack, transcriptAnnotation="symbol")
## Two miRNAs are encoded in the first exon of SKA2.
```

```{r other-packages-mirhostgenes}
## Load a database defining predicted miRNA host genes for miRBase 21.
library(mirhostgenes)
library(MirhostDb.Hsapiens.v81.v21)
mhdb <- MirhostDb.Hsapiens.v81.v21

## Re-use the GenenameFilter
column(gnf, mhdb)
where(gnf, mhdb)

## Get all mature miRNAs that could be encoded in a common primary transcript.
matmirnas(mhdb, filter=gnf)
```

-   miRNA target genes: `mirtarbase` package
    (<https://github.com/jotsetung/mirtarbase>), bases on data from miRTarBase
    (<http://mirtarbase.mbc.nctu.edu.tw/>).
-   All miRNA-target gene interactions (MTIs) for all species derived from
    literature.
-   Imports all filter classes from `ensembldb` and `mirhostgenes`.
-   <span class="underline">Example</span>: get all miRNAs targeting SKA2 and all target genes of miR-301a-3p.

```{r }
####
## Load the mirtarbase library providing validated target genes
library(mirtarbase)

## Check wat we've got.
mirtarbase

## Target genes are categorized by 'Support type'
listSupportTypes(mirtarbase)

## Re-use the GenenameFilter
column(gnf, mirtarbase)
where(gnf, mirtarbase)

## Get all miRNA-target gene interactions for SKA2.
miRs <- mtis(mirtarbase, filter=gnf)
miRs

## What miRNAs have been shown to regulated SKA2?
matmirna(miRs)

####
## Get all miRNA-target gene interactions (MTIs) for miR-301a-3p.
tarGenes <- mtis(mirtarbase, filter=MatmirnaFilter("hsa-miR-301a-3p"))
tarGenes

## Restrict to highest confidence.
tarGenes <- mtis(mirtarbase, filter=list(MatmirnaFilter("hsa-miR-301a-3p"),
					 SupportTypeFilter("Functional MTI")))
tarGenes

## Get the genes.
gene(tarGenes)

## Use this back in EnsDb
genes(edb, filter=GenenameFilter(gene(tarGenes)))
```

-   Both packages are still in development.

## Future stuff<a id="orgheadline16"></a>

-   Combine filters similar to the `Rules` objects: this would need `FilterSet`
      and `FilterSetList` objects.
